Length Bias and Search Limitations in Cartesian Genetic Programming

### 2013-04-10 Length Bias and Seach Limitations in CGP/main.tex
Minor fixes

\section{Introduction}
In all black box search, search designers must make assumptions about the properties
of the problem in order to outperform random search.  When these assumptions
match the application, search time can be greatly reduced.  As such, understanding how design decisions
effect search effectiveness or even prevent search from finding solutions is very important.
%
In Cartesian Genetic Programming (CGP) the maximum genome size of an individual
is set at initialization.  While this
requires the maximum genome size to be larger than the problem's solution needs,
the question remains; how will the maximum genome size effect evolution's
speed at finding solutions.  As larger genomes inherently represent larger search
spaces, intuition suggests that using the lowest possible maximum size will have the most efficient search.
However, in CGP a mounting body of evidence suggests the optimal
maximum size is much larger than the actual solutions evolved~\cite{miller:2006:redundancy}.
We examine the cause of this discrepancy by applying CGP and a few
modified versions of CGP to problems which test specific edge cases in representation.
The result is a better understanding of the search limitations of CGP.

\section{CGP}
\label{sec-cgp}
In Cartesian Genetic Programming (CGP) individuals use a fixed length string of integers to
encode directed acyclic graphs (DAGs).  
Each node in the DAG is encoded using multiple genes: one specifying a function and
the rest specifying node input connections.
Graph output is determined by additional genes which can connect to any node or
program input.
To prevent cycles,
nodes may only connect to problem inputs or nodes that precede them in the genome.
Early versions of CGP limited valid connections further through user specified graph
topologies.
While specifying topology is still useful
when evolving circuits with known hardware limitations, the less restricted form
allows CGP to represent all possible
DAGs of size \emph{N}, where \emph{N} is the number of nodes in the genome.
%
The most common evolutionary mechanism in CGP is a $1+\lambda$ population
where each generation the best offspring replaces the parent if it is no less
fit.  CGP almost exclusively uses a form of point mutation to create offspring,
with no form of crossover commonly being employed.  As such, these are the methods
considered in our analysis.
For a more detailed description of CGP, see~\cite{miller:2011:chapter2}.

@@

When initializing and mutating a connection gene, its value is randomly set to
either a preceding node in the genome or to an input.  As was discussed in~\cite{collins:2006:haystacks},
this means the expected location of a new connection is $\frac{i}{2}$ where \emph{i}
is the number of nodes and inputs that precede the mutated node.  If all nodes had
a single connection gene, we would expect random individuals to have a length of $\log_2N$.
Precise calculation of the expected length is made difficult
due to node overlap, but the relationship will remain logarithmic to the total genome size.

@@

The second experiment looked at the distribution of evolved lengths using a fixed
genome size of 100. This
information is shown in Figure~\ref{fig-length} as a probability density curve,
with numeric results in Table~\ref{table-length}.
The results for \emph{Normal} match our predictions in Section~\ref{sec-reorder};
most individuals were highly inactive, with a mode number of
active nodes of only 15.  The probability of generating an individual drops quickly
as the number of active nodes deviates from the mode, with none of the 0.5 billion
generated individuals having more than 59 active nodes, and only one in 10,000 above 45.  This is in contrast
to the actual number of graphs of each size, which increases exponentially.
This means that \emph{Normal} is biased toward small graphs even more than
Figure~\ref{fig-length} indicates.
\emph{Reorder} and \emph{DAG} show less of a bias, with \emph{DAG}'s mode of 45
close to the prediction made in Section~\ref{sec-dag}.  The implications of these
biases is that each is most likely to find solutions fastest when the number of
active nodes desired by the problem best matches the algorithms search bias, as
otherwise the algorithm will add artificial pressure to search different sized
graphs.  The steepness of the bias reflects how much pressure will be applied.
Combined with the length scaling results in Figure~\ref{fig-neutral}, we would
predict \emph{Normal} to both require a larger genome size than \emph{Reorder} and \emph{DAG},
as well as search a much smaller subset of the search space.

>>>>

\section{Introduction}
In all black box search, search designers must make assumptions about the properties
of the problem in order to outperform random search.  When these assumptions
match the application, search time can be greatly reduced.  As such, understanding how design decisions
effect search effectiveness or even prevent search from finding solutions is very important.
%
In Cartesian Genetic Programming (CGP) the maximum genome size of an individual
is set at initialization.  While this
requires the maximum genome size to be larger than the problem's solution needs,
the question remains; how will the maximum genome size effect evolution's
speed at finding solutions?  As larger genomes inherently represent larger search
spaces, intuition suggests that using the lowest possible maximum size will have the most efficient search.
However, in CGP a mounting body of evidence suggests the optimal
maximum size is much larger than the actual solutions evolved~\cite{miller:2006:redundancy}.
We examine the cause of this discrepancy by applying CGP and a few
modified versions of CGP to problems which test specific edge cases in representation.
The result is a better understanding of the search limitations of CGP.

\section{CGP}
\label{sec-cgp}
In Cartesian Genetic Programming (CGP) individuals use a fixed length string of integers to
encode directed acyclic graphs (DAGs).  
Each node in the DAG is encoded using multiple genes: one specifying a function and
the rest specifying node input connections.
Graph output is determined by additional genes which can connect to any node or
program input.
To prevent cycles,
nodes may only connect to problem inputs or nodes that precede them in the genome.
Early versions of CGP limited valid connections further through user specified graph
topologies.
While specifying topology is still useful
when evolving circuits with known hardware limitations, the less restricted form
allows CGP to represent all possible
DAGs of size \emph{N}, where \emph{N} is the number of nodes in the genome.

%
The most common evolutionary mechanism in CGP is a $1+\lambda$ population
where each generation the best offspring replaces the parent if it is no less
fit.  CGP almost exclusively uses a form of point mutation to create offspring,
with no form of crossover commonly being employed.  As such, these are the methods
considered in our analysis.
For a more detailed description of CGP, see~\cite{miller:2011:chapter2}.

@@

When initializing and mutating a connection gene, its value is randomly set to
either a preceding node in the genome or to an input.  As was discussed in~\cite{collins:2006:haystacks},
this means the expected location of a new connection is $\frac{i}{2}$ where \emph{i}
is the number of nodes and inputs that precede the mutated node.  If all nodes had
a single connection gene, we would expect random individuals to have a length of $\log_2N$.
Precise calculation of the expected length when using more connection genes is made difficult
due to node overlap, but the relationship will remain logarithmic to the total genome size.

@@

The second experiment looked at the distribution of evolved lengths using a fixed
genome size of 100. This
information is shown in Figure~\ref{fig-length} as a probability density curve,
with numeric results in Table~\ref{table-length}.
The results for \emph{Normal} match our predictions in Section~\ref{sec-reorder};
most individuals were highly inactive, with a mode number of
active nodes of only 15.  The probability of generating an individual drops quickly
as the number of active nodes deviates from the mode, with none of the 0.5 billion
generated individuals having more than 59 active nodes, and only one in 10,000 above 45.  This is in contrast
to the actual number of graphs of each size, which increases exponentially.
This means that \emph{Normal} is biased toward small graphs even more than
Figure~\ref{fig-length} indicates.
\emph{Reorder} and \emph{DAG} show less of a bias, with \emph{DAG}'s mode of 45
close to the prediction made in Section~\ref{sec-dag}.  The implications of these
biases is that each is most likely to find solutions fastest when the number of
active nodes desired by the problem best matches the algorithm's search bias, as
otherwise the algorithm will add artificial pressure to search different sized
graphs.  The steepness of the bias reflects how much pressure will be applied.
Combined with the length scaling results in Figure~\ref{fig-neutral}, we would
predict \emph{Normal} to both require a larger genome size than \emph{Reorder} and \emph{DAG},
as well as search a much smaller subset of the search space.
